Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2021-12-15 analysis/paper/paper.Rmd
EDITS THROUGH METHODS
initial edits in response to Justin C comments

Ice-associated seals rely on sea ice for a variety of activities, including
breeding, molting, pupping, and resting. In the Arctic, many of these activities
occur in spring (April-June) as sea ice begins to melt and retreat northward.
Rapid acceleration of climate change in Arctic ecosystems is therefore of
concern as the quantity and quality of suitable habitat is forecast to decrease.
In this paper, we use data collected from satellite-linked bio-loggers deployed
between 2005 and 2020 to investigate the seasonal timing and environmental
factors affecting haul-out behavior sea ice by seals (specifically, bearded,
ribbon, and spotted seals) in the Bering and Chukchi seas. In addition to
providing baseline data on phenology, these data also allow us to quantify
'availability', which is needed to accurately estimate abundance from aerial
survey counts of seals basking on ice (i.e., to correct for the proportion of
animals that are in the water while surveys are conducted). Using generalized
linear mixed pseudo-models to properly account for temporal autocorrelation, we
fit models with a variety of covariates (e.g., day-of-year, solar hour, age-sex
class, wind speed, barometric pressure, temperature, precipitation) to examine
their ability to explain variation in hourly haul-out records. We found evidence
for strong diurnal and within-season patterns in haul-out behavior, as well as
strong weather effects (particularly wind and temperature). In general, seals
were more likely to haul out on ice in the middle of the day and when wind speed
was low and temperatures were higher. Haul-out propensity increased through
March and April, peaking in May and early June before declining again. The
timing and frequency of haul-out events also varied based on species and age-sex
class. For ribbon and spotted seals, models with year effects were highly
supported, indicating that the timing and magnitude of haul-out behavior varied
among years. However, we did not find evidence that haul-out timing was linked
to annual sea ice extent. Our analysis emphasizes the importance of accounting
for variation in haul-out behavior when interpreting the number of seals counted
in aerial surveys.

@@

The global climate disruption is causing considerable reduction in Arctic sea
ice extent, volume, and seasonal presence [@kwok2018;
@meier2014; @overland2021; @wang2017]. These changes have ripple effects on
Arctic organisms, ecosystems, and the human communities who live in the region
[@huntington2020]. Such disruptions are a particular cause of concern for
ice-associated seals (ringed (*Pusa hispida*), bearded (*Erignathus barbatus*),
ribbon seals (*Histriophoca fasciata*), and spotted (*Phoca largha*)) that
depend on spring and early summer sea ice (March-June) in the Bering and Chukchi
seas as a platform for important functions, such as pupping, nursing, breeding,
and molting [@boveng2009; @boveng2013a; @cameron2010; @kelly2010]. Limited data
and large knowledge gaps complicate predictions about the ultimate effects of
changes in sea ice on the behavior, abundance, and distribution of these seals.
Knowledge about evolutionary constraints on the timing of reproductive and
molting behavior is generally lacking, so it is difficult to predict the
readiness with which ice-associated seal species will be able to adapt to future
changes (e.g., by adjusting pupping or molting schedules to earlier dates or
different locales). Additionally, trends in abundance of these species are
largely unknown, so it is difficult to say what effect, if any, declines in sea
ice habitat have had, or will have, on seal densities.

For ice-associated seals, haul-out behavior is tightly linked with sea ice.
Ribbon seals haul out of the water almost exclusively on sea ice and are mostly
pelagic outside the spring pupping, breeding, and molting season [@boveng2018a].
While spotted and bearded seals rest on beaches during ice-free months, they
strongly prefer sea ice as a resting platform during the spring and early summer
[@frost2018]. The remoteness of the Bering and Chukchi seas mean direct
scientific observation of seal behavior is impractical. Thus, bio-logging
devices are, thus, especially useful tools for collecting key information on
movement and haul-out behavior for these species.

@@

While analyzing haul-out patterns clearly has utility for understanding natural
history and ecology, some studies have focused, as we do here, specifically on
quantifying the probability of seals being hauled out as a means of quantifying
seals in the water that are not counted during aerial surveys. These studies
typically use logistic regression-style analyses to estimate the time-specific
probability of being hauled out based on 'wet/dry' data relayed by bio-loggers.
In these models, haul-out probabilities were expressed as a function of
predictive covariates, such as time of day, day-of-year, sex, age class, and
environmental conditions (e.g., [@reder2003], [@bengtson2004], [@bengtson2005a],
[@udevitz2009], [@verhoef2014a], and [@southwell2008]). However, sample sizes
have often been insufficient to permit strong inference about life history
and/or seasonal variation in haul-out probabilities. For instance, Bengtson and
Cameron's [-@bengtson2004] study included 5 adult and 2 juvenile crabeater
seals, while Bengtson et al.'s [-@bengtson2005a] study was based on 6 ringed
seals in the Chukchi Sea. These studies were often further limited by logistical
constraints on tagging fieldwork and the attachment duration or operational life
of bio-loggers. In this study, we address some of these limitations by deploying
small bio-loggers designed for longer-term attachment on a subset of the study
individuals. These devices are designed to collect data through the molt period
(when those adhered to the hair would fall off) and, in some situations, provide
multiple years of data.

Ultimately, knowledge of trends in phenology and abundance (or life history
surrogates such as survival and recruitment) will be necessary to make credible
quantitative predictions about the effects of climate disruptions on the
abundance and distribution of Arctic seal populations. Before we can construct a
trend, however, we first require a baseline. Several studies have contributed
estimates of the distribution and abundance of ice-associated seal species in
the Arctic using aerial surveys (e.g., [@bengtson2005a], [@conn2014a], and
[@verhoef2014a]). Such studies were conducted over very large areas and
estimation of absolute abundance required making inference about numerous issues
affecting the observation of seals on ice. These included availability (only
seals basking on ice were available to be counted), detection probability
(observers or automated detection systems may have missed some seals on ice),
species misclassification, and possible disturbance of seals by aircraft
[@conn2014a; @verhoef2014a].

In this study, we used 16 years of bio-logging data to investigate the haul-out
behavior of 3 species of ice-associated seals in the Bering and Chukchi seas. We
focused our effort on bearded, ribbon, and spotted seals and did not include
ringed seals in the analysis. Ringed seals are unique in their use of snow lairs
in the spring months, and they warrant a separate study that includes additional
environmental covariates, such as the onset of snow melt and snow depth. Our
goals are threefold. First, we wish to establish baseline estimates for the
chronology of haul-out behavior in the critical spring season for each species
across different age and sex classes. Second, we wish to refine estimates of
haul-out availability corrections for aerial surveys in order to improve
estimates of seal abundance. Previously estimated availability correction
factors (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a]) accounted
for variables such as the time-of-day and day-of-year, but did not investigate
meteorological variables that have been shown to influence haul-out behavior of
walruses [@reder2003; @udevitz2009]. Third, we explored the annual variability
in haul-out timing and possible linkage to changes in the extent of seasonal sea
ice between 2005 and 2020. Our work extends the scope of previous haul-out
analyses, includes the influence of meteorological variability, and investigates
the potential impact of a changing icescape on the behavior of these species.

@@

```{r getdata}
targets::tar_load(analysis_data, store = here::here("_targets"))
targets::tar_load(grid, store = here::here("_targets"))
targets::tar_load(deploy_table, store = here::here("_targets"))
```

Haul-out behavior data from `r dplyr::n_distinct(deploy_table$speno)`
bio-loggers deployed on bearded, ribbon, and spotted seals was filtered to
include only records from 1 March to 15 July between `r
min(lubridate::year(analysis_data$haulout_dt))` and `r
max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA) and either adhered to the hair on the seal or attached to the
rear flipper inter-digital webbing. The use of bio-loggers adhered to the back
or head provide some benefits over flipper mounted devices (e.g. increased
satellite transmittal rates, locations at sea) but these are lost during the
annual molt, which limits the effective length of haul-out timelines.
Additionally, bio-loggers attached to the head or dorsal region are often dry
while the seal is at the surface and can slightly bias the percent-dry values
recorded by the device. For this study, in cases where both tag types were
deployed, percent dry observations from the flipper tag were preferred. Sex and
age class (non-dependent *young-of-the-year*, sexually immature *subadults*, and
mature *adults*) were determined at the time of deployment by varying
combinations of length, claw annuli, and pelage characteristics for some
species. Field identification of age class can be inexact, particularly when
discerning subadults from adults. In the case of ribbon seals, pelage
differences between subadults and adults are distinguishable, with subadults
often having less defined ribbons. Bearded seal subadults will often have a
spotted pattern in the pelage that is not seen in adults. Spotted seal pelage is
monomorphic and can not be used to discern age class. Despite these challenges,
we feel the age classifications used in this analysis are useful categories in
testing for age-related effects on haul-out behavior. Seals determined to be
less than one year were classified as young-of-the-year. Subadults are those
seals likely greater than one year of age but not yet sexually mature. Adults
are likely sexually mature and older than four years. For those
bio-loggers deployed on young-of-the-year and transmitting into the next year,
the age class was advanced to subadult on 1 March of the following year. Table
\@ref(tab:tblDeploy) provides a summary of these deployments and data received
from them.

@@

Tags that fall off due to molt or mortality and remain on ice or land can still
send data to satellites, and we did not want to include data from these tags in
our analyses. As such, end times of each deployment were identified by examining
tag locations and dive behavior to determine when tags ceased providing data
consistent with seal behavior. Sensor readings outside of these start and end
times were discarded prior to analysis. After approximately 9 months, 7 devices
deployed on the rear flipper of bearded seals reported implausible sensor data.
All data after the first instance of unrealistic values were censored from this
analysis. Figure \@ref(fig:dataCal) shows a calendar-like distribution of all
haul-out data across the study season for each species. Observations for ribbon
and spotted seals are concentrated in the months of May and June due to the
timing of deployment (April and May) and the timing of molt (May and June).
During molt, seals (and their attached bio-loggers) spend more time out of the
water and more data is transmitted. Molt timing also impacts when many
deployments end as any tags adhered to the hair will fall off. Bearded seal
deployments are more evenly spread with fewer observations in May and June. The
majority of bearded seal deployments start later in the summer and by May tags
have either fallen off or the batteries are depleted.

@@

Haul-out behavior data was recorded and transmitted via the Argos satellite
network as hourly percent-dry timelines. For each hour of a day, the wet/dry
sensor was polled by the tag firmware every few seconds and a percent of the
hour in a dry state was calculated (Figure \@ref(fig:examplePlot)). Each data
transmission message consisted of a complete 24-hour record. Bio-loggers were,
generally, deployed and programmed in a manner to maximize data transmission
during the spring pupping and molting period, however, not all days during a
deployment were successfully transmitted. This is due to a variety of factors
including satellite coverage, tag availability (i.e. tags mounted to the rear
flipper are not available to the surface while the seal is at sea), tag
performance, duty cycling, and extra-terrestrial interference.

@@

Of key interest in this study is the relationship between haul-out behavior and
environmental covariates that vary with space and time. We explored
the use of a movement model to predict locations at specific times. However, the
sparse nature of data from some tags, especially those mounted to the rear
flipper, resulted in poor modeling performance or convergence issues. For this
study, we calculated a weighted average daily location where the inverse of the
estimated Argos location error was used for the weight. Any days where haul-out
observations were present without any location data were filled from the last
calculated weighted average daily location, and any days where the location
intersected with land were removed from the dataset. Figure \@ref(fig:dataMap)
shows the spatial distribution of determined locations and available haul-out
behavior data for each species across the study area.

@@

In addition to sex and age class, we analyzed variables that might
help explain variation in haul-out probabilities. These included day-of-year
(for seasonal effects) and local solar hour (for diurnal effects). Solar hour
was calculated using the {solaR} package [@perpinan2012] within the R
statistical environment [@rcoreteam2021] using the daily locations. We included
several meteorological variables shown to affect haul-out behavior in
other Arctic pinnipeds [@reder2003; @udevitz2009; @perry2017]. In particular, we
linked locations from our biologgers to key weather values from the North
American Regional Reanalysis (NARR) model produced by the National Centers for
Environmental Prediction [@mesinger2006]. The NARR model assimilates
observational data to produce a long-term picture of weather over North America.
Numerous weather variables are made available across the region 8 times daily.
For this study, NARR weather values were subset to the extent of our study area
over the Bering and Chukchi seas at 3-hr intervals based on the native grid
resolution of 32 km (1024 sq. km). The following meteorological variables were
interpolated and assigned to seal locations using a bilinear method: 1) air
temperature at 2m above the earth's surface, 2) wind consisting of northerly and
easterly vector components converted to wind speed using the Euclidean norm, 3)
barometric pressure at sea level, and 4) precipitation (Table
\@ref(tab:tblCovariates)).

Models for all species include the following effects: day-of-year, solar hour,
temperature, wind speed, barometric pressure, precipitation, wind chill
(represented by a *wind:temperature* interaction [@udevitz2009]), and
day-of-year and time of day interactions to permit diurnal patterns to change
throughout the year. Ribbon and spotted seal models included age-sex class and
interactions between day-of-year and age-sex class, but we omitted these from
bearded seal models due to poor representation of age-sex classes (Table
\@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect (and an
interaction with day-of-year), since bearded seals occupy a substantial range
and we were interested in possible differences in the timing of haul-out
behavior along a latitudinal gradient. We omitted the latitudinal effect from
ribbon and spotted seal models since, during the spring, these species are most
prevalent near the southern ice edge in the Bering Sea [@conn2014a].

Notably missing from this list of explanatory variables is any spatial-temporal
representation of sea ice concentration, area, or extent. This may seem
counterintuitive when modeling the haul-out behavior of seal species with such a
close association to sea ice; seals haul out in the presence of sea ice, and we
could assess the local concentration of sea ice during these events (see
[@olnes2020]). However, a major focus of this study is to develop models
applicable for aerial survey correction factors and using sea ice as a covariate
would almost certainly bias haul-out predictions towards those seals that are on
or near ice and therefore more likely to haul out. Since aerial surveys can only
detect seals on ice, abundance estimates would be missing a correction for those
seals that are away from ice (e.g., on foraging trips). Additionally, many of
the bio-logger deployments consisted of a single device attached to the
rear flipper of the seal which do not provide at-sea locations, limiting our
ability to fully evaluate habitat preferences related to sea-ice. It should be
noted that the aerial survey counts are modeled with sea-ice and other habitat
parameters. So, any final abundance estimates that combine models of survey
counts and our model for proportion of seals hauled out on ice do account for
the strong association between these species and sea-ice in the Bering and
Chukchi seas. Lastly, our study was limited to the spring season when seal
haul-out tendencies are strongly influenced by pupping, nursing, breeding behavior, and
molt and these drivers are likely more influential than specific sea-ice
concentration. Crawford et al [-@crawford2019] compared haul-out probability
models for ringed seals and found those that only included season (and not
sea-ice concentration) were the most parsimonious. For these reasons, we have
elected not to use sea ice concentration as a predictor for haul-out
probability; studies to evaluate habitat selection in relation to sea-ice
parameters and other environmental covariates are in progress.

We explored whether the annual variation in maximum spring sea ice extent in the
Bering Sea influences the seasonal peak of seal haul-out behavior. In
particular, we used sea ice concentration data from the Nimbus-7 SMMR and DMSP
SSM/I-SSMIS Passive Microwave Dataset, Version 1 [@cavalieri1996] to calculate
maximum sea ice extent. All sea ice concentration grid cells (25 km^2^) in the
study area with greater than 15% concentration were summed to get the total sea
ice extent for each day between 15 February and 15 July across all years.
Maximum spring sea ice extent was simply the largest daily value for each year.

@@

We followed Ver Hoef et al. [-@verhoef2014a] in using linear, quadratic, and
cubic effects of day-of-year to represent seasonal changes in behavior. However,
unlike previous models for harbor seals [@london2012] and ice-associated seals
[@verhoef2014a], which treated hour-of-day as a 24-level categorical variable to
capture diurnal cycles, we adopted a continuous formulation based on Fourier
series that provides a flexible model while preserving the inherent circularity
needed for time-of-day effects (i.e., hour 0 should be equal to hour 24). It
also represents hour-of-day with 6 parameters, which is a considerable reduction
when compared to a 24 parameter variable, especially when fitting models with
interactions between hour-of-day and other variables (e.g., age-sex class,
day-of-year). According to this approach, we used the following specification
for hour-of-day effects:

>>>>

Ice-associated seals rely on sea ice for a variety of activities, including
breeding, molting, pupping, and resting. In the Arctic, many of these activities
occur in spring (April-June) as sea ice begins to melt and retreat northward.
Rapid acceleration of climate change in Arctic ecosystems is therefore of
concern as the quantity and quality of suitable habitat is forecast to decrease.
In this paper, we use data collected from satellite-linked bio-loggers deployed
between 2005 and 2020 to investigate the seasonal timing and environmental
factors affecting haul-out behavior by ice-associated seals (specifically, bearded,
ribbon, and spotted seals) in the Bering and Chukchi seas. In addition to
providing baseline data on phenology, these data also allow us to quantify
'availability', which is needed to accurately estimate abundance from aerial
survey counts of seals basking on ice (i.e., to correct for the proportion of
animals that are in the water while surveys are conducted). Using generalized
linear mixed pseudo-models to properly account for temporal autocorrelation, we
fit models with covariates of interest (e.g., day-of-year, solar hour, age-sex
class, wind speed, barometric pressure, temperature, precipitation) to examine
their ability to explain variation in hourly haul-out probability. We found evidence
for strong diurnal and within-season patterns in haul-out behavior, as well as
strong weather effects (particularly wind and temperature). In general, seals
were more likely to haul out on ice in the middle of the day and when wind speed
was low and temperatures were higher. Haul-out probability increased through
March and April, peaking in May and early June before declining again. The
timing and frequency of haul-out events also varied based on species and age-sex
class. For ribbon and spotted seals, models with year effects were highly
supported, indicating that the timing and magnitude of haul-out behavior varied
among years. However, we did not find evidence that haul-out timing was linked
to annual sea ice extent. Our analysis emphasizes the importance of accounting
for seasonal and temporal variation in haul-out behavior, as well as associated
environmental covariates, when interpreting the number of seals counted
in aerial surveys.

@@

The global climate disruption is causing considerable reduction in Arctic sea
ice extent, volume, and seasonal presence [@kwok2018;
@meier2014; @overland2021; @wang2017]. These changes have ripple effects on
Arctic organisms, ecosystems, and the human communities who live in the region
[@huntington2020]. Such disruptions are a particular cause of concern for
ice-associated seals (ringed (*Pusa hispida*), bearded (*Erignathus barbatus*),
ribbon seals (*Histriophoca fasciata*), and spotted (*Phoca largha*)) that
depend on spring and early summer sea ice (March-June) in the Bering and Chukchi
seas as a platform for important functions, such as pupping, nursing, breeding,
and molting [@boveng2009; @boveng2013a; @cameron2010; @kelly2010]. Limited data
and large knowledge gaps complicate predictions about the ultimate effects of
changes in sea ice on the behavior, health, abundance, and distribution of these
seals. To date, indices of seal health sampled during periods of declining sea
ice differ regionally [@crawford2015; @harwood2020]. Knowledge about
evolutionary constraints on the timing of reproductive and molting behavior is
generally lacking, so it is difficult to predict the readiness with which
ice-associated seal species will be able to adapt to future changes (e.g., by
adjusting pupping or molting schedules to earlier dates or different locales).
Additionally, trends in abundance of these species are largely unknown, so it is
difficult to assess the effect, if any, declines in sea ice habitat have had, or
will have, on seal densities.

For ice-associated seals, haul-out behavior is tightly linked with sea ice.
Ribbon seals haul out of the water almost exclusively on sea ice and are mostly
pelagic outside the spring pupping, breeding, and molting season [@boveng2018a].
While spotted and bearded seals rest on coastal features, they
strongly prefer sea ice as a resting platform during the spring and early summer
[@frost2018]. The remoteness of the Bering and Chukchi seas mean direct
scientific observation of seal behavior is impractical. Thus, bio-logging
devices are, thus, especially useful tools for collecting key information on
movement and haul-out behavior for these species.

@@

While analyzing haul-out patterns clearly has utility for understanding natural
history and ecology, some studies have focused, as we do here, specifically on
quantifying the probability of seals being hauled out as a means of quantifying
seals in the water that are not counted during aerial surveys. These studies
typically use logistic regression-style analyses to estimate the time-specific
probability of being hauled out based on 'wet/dry' data relayed by bio-loggers.
In these models, haul-out probabilities were expressed as a function of
predictive covariates, such as time-of-day, day-of-year, sex, age class, and
environmental conditions (e.g., [@reder2003], [@bengtson2004], [@bengtson2005a],
[@udevitz2009], [@verhoef2014a], and [@southwell2008]). However, sample sizes
have often been insufficient to permit strong inference about life history
and/or seasonal variation in haul-out probabilities. For instance, Bengtson and
Cameron's [-@bengtson2004] study included 5 adult and 2 juvenile crabeater
seals, while Bengtson et al.'s [-@bengtson2005a] study was based on 6 ringed
seals in the Chukchi Sea. These studies were often further limited by logistical
constraints on fieldwork and the attachment duration or operational life
of bio-loggers. In this study, we address some of these limitations by deploying
small bio-loggers designed for longer-term attachment on a subset of the study
individuals. These devices are designed to collect data through the molt period
(when those adhered to the hair would fall off) and, in some situations, provide
multiple years of data.

Ultimately, knowledge of trends in phenology and abundance (or life history
surrogates such as survival and recruitment) will be necessary to make credible
quantitative predictions about the effects of climate disruptions on the
abundance and distribution of Arctic seal populations. Before we can construct a
trend, however, we first require a baseline. Several studies have contributed
estimates of the distribution and abundance of ice-associated seal species in
the Arctic using aerial surveys (e.g., [@bengtson2005a], [@conn2014a], and
[@verhoef2014a]). Such abundance studies were conducted over very large areas and
estimation of absolute abundance required making inference about numerous issues
affecting the observation of seals on ice. These included availability (only
seals basking on ice were available to be counted), detection probability
(observers or automated detection systems may have missed some seals on ice),
species misclassification, and possible disturbance of seals by aircraft
[@conn2014a; @verhoef2014a]. Refining these inferences will improve the
accuracy of abundance estimates in the Arctic.

In this study, we used 16 years of bio-logging data to investigate the haul-out
behavior of 3 species of ice-associated seals in the Bering and Chukchi seas. We
focused our effort on bearded, ribbon, and spotted seals and did not include
ringed seals in the analysis. Ringed seals are unique in their use of snow lairs
in the spring months, and they warrant a separate study that includes additional
environmental covariates, such as the onset of snow melt and snow depth. Our
goals were threefold. First, we wish to establish baseline estimates for the
chronology of haul-out behavior in the critical spring season for each species
across different age and sex classes. Second, we refined estimates of
haul-out availability corrections for aerial surveys in order to improve
estimates of seal abundance. Previously estimated availability correction
factors (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a]) accounted
for variables such as the time-of-day and day-of-year, but did not investigate
meteorological variables that have been shown to influence haul-out behavior of
walruses [@reder2003; @udevitz2009]. Third, we assessed the annual variability
in haul-out timing and possible linkage to changes in the extent of seasonal sea
ice between 2005 and 2020. Our work extends the scope of previous haul-out
analyses, includes the influence of meteorological variability, and investigates
the potential impact of a changing icescape on the behavior of these species.

@@

```{r getdata}
targets::tar_load(analysis_data, store = here::here("_targets"))
targets::tar_load(grid, store = here::here("_targets"))
targets::tar_load(deploy_table, store = here::here("_targets"))
```
Data used in our study came from bio-loggers deployed on bearded, ribbon, and
spotted seals in the Bering, Chukchi, and Beaufort seas by multiple
organizations as part of collaborative investigations during 2005 and 2020.
Seals were captured using nets and instrumented during ship-based capture trips
and from coastal Alaska communities. Ship-based capture events occurred during
spring near the southern ice edge in the Bering Sea from 20xx to 20xx (maybe
provide sample sizes for each species). Land-based capture events occurred from
May to October, generally between the coastal communities of Scammon Bay,
Alaska, in the Bering Sea, and Utqiaġvik, Alaska, near the junction of the
Chukchi and Beaufort seas, from 2005 to 2020. We refer readers to the primary
literature for detailed capture and bio-logger attachment methods (Supplemental
Material, Table).

Haul-out behavior data from `r dplyr::n_distinct(deploy_table$speno)`
bio-loggers deployed on bearded, ribbon, and spotted seals were subset to
include only records from 1 March to 15 July between `r
min(lubridate::year(analysis_data$haulout_dt))` and `r
max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA) and either adhered to the hair on the seal or attached to the
rear flipper inter-digital webbing. The use of bio-loggers adhered to the back
or head provide some benefits over flipper mounted devices (e.g. increased
satellite transmittal rates, locations at sea) but these are lost during the
annual molt, which limits the duration of haul-out data they provide.
Additionally, bio-loggers attached to the head or dorsal region are often dry
while the seal is at the surface and can slightly bias the hourly percent-dry values
reported by the bio-logger. For this study, in cases where both bio-logger types were
deployed, hourly percent dry observations from the flipper tag were preferred. Sex and
age class (non-dependent *young-of-the-year*, sexually immature *subadults*, and
mature *adults*) were determined at the time of deployment by varying
combinations of length, claw annuli [@burns1969; @burns1983; @mclaren1958a], and pelage characteristics for some
species. Field identification of age class can be inexact, particularly when
discerning subadults from adults. In the case of ribbon seals, pelage
differences between subadults and adults are distinguishable, with subadults
often having less defined ribbons. Bearded seal subadults will often have a
spotted pattern in the pelage that is not seen in adults. Spotted seal pelage is
monomorphic and can not be used to discern age class. Despite these challenges,
we feel the age classifications used in this analysis are useful categories in
testing for age-related effects on haul-out behavior. Seals determined to be
less than one year were classified as young-of-the-year. Subadults are those
seals likely greater than one year of age but not yet sexually mature. Adults
are likely sexually mature and older than four years. For those
bio-loggers deployed on young-of-the-year and transmitting into the next year,
the age class was advanced to subadult on 1 March of the following year. Table
\@ref(tab:tblDeploy) provides a summary of these deployments and data received
from them.

@@

Tags that fall off due to molt or mortality and remain on ice or land can still
send data to satellites; i.e., the bio-logger will be on ice or land and dry, therefore, it will record and transmit data suggesting the seal is hauled out. As such, end times of each deployment were identified by examining
bio-logger locations and dive behavior to determine when bio-loggers ceased providing data
consistent with seal behavior. Data outside of the deployment start and end
times were discarded prior to analysis. After approximately 9 months, 7 devices
deployed on the rear flipper of bearded seals reported implausible hourly percent-dry data.
All data after the first instance of unrealistic values were censored from this
analysis. Figure \@ref(fig:dataCal) shows a calendar-like distribution of all
haul-out data across the study season for each species. Observations for ribbon
and spotted seals are concentrated in the months of May and June due to the
timing of deployment (April and May) and the timing of molt (May and June).
During molt, seals (and their attached bio-loggers) spend more time out of the
water and more data are transmitted. Molt timing also impacts when many
deployments end as any bio-loggers adhered to the hair will fall off. Bearded seal
deployments are more evenly spread with fewer observations in May and June. The
majority of bearded seal deployments start later in the summer and by May bio-loggers
have either fallen off or the batteries are depleted.

@@

Haul-out behavior data was recorded and transmitted via the Argos satellite
network as hourly percent-dry timelines. For each hour of a day, the wet/dry
sensor was polled by the tag firmware every few seconds and a percent of the
hour in a dry state was calculated (Figure \@ref(fig:examplePlot)). Hourly
percent-dry data were usually rounded to the nearest 10 percent inclusive of 0
and 100 percent and additional values at 3 and 98 percent. This compression
resulted in additional data transmission as each message consisted of two
complete 24-hour records. Bio-loggers were, generally, deployed and programmed
in a manner to maximize data transmission during the spring pupping and molting
period, however, hourly percent-dry data for days during a deployment were not
successfully transmitted. This is due to a variety of factors including
satellite coverage, tag availability (i.e. tags mounted to the rear flipper are
not available to the surface while the seal is at sea), tag performance, duty
cycling, and extra-terrestrial atmospheric interference.

@@

Of key interest in this study is the relationship between haul-out behavior and
environmental covariates that vary with space and time. We explored the use of a
continuous-time correlated random walk [@devins.johnson2008] movement model to
predict locations at specific times. However, the sparse nature of data from
some bio-loggers, especially those mounted to the rear flipper, resulted in poor
modeling performance or convergence issues. For this study, we calculated a
weighted average daily location where the inverse of the estimated Argos or
FastLoc GPS location error was used for the weight. Each Argos location estimate
was assigned an error radius based on either the categorical location quality (
_3_=250m, _2_=500m, _1_=1500m, _0_=2500m, _A_=2500m, _B_=2500m) or, when
available, the estimated error radius from the Argos Kalman filter algorithm.
Location estimates from FastLoc GPS were all assigned an error radius of 50m.
Any days where haul-out observations were present but location data were missing
we used the last calculated weighted average daily location, and any days where
the location intersected with land were removed from the data set. We recognized
that bearded and spotted seals haul out on land. However, assessing the
relationship between haul-out behavior and environmental covariates and their
availability for aerial surveys on land was outside the scope of this study.
Additionally, any daily locations on land are likely more reflective of
coordinate averaging and measurement error instead of actual use of coastal
features. Figure \@ref(fig:dataMap) shows the spatial distribution weighted
locations with available haul-out behavior data for each species across the
study area.

@@

In addition to sex and age class, we analyzed variables that might help explain
variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and local solar hour (for diurnal effects). Solar hour was calculated
using the {solaR} package [@perpinan2012] within the R statistical environment
[@rcoreteam2021] using the daily locations. We included several meteorological
variables shown to affect haul-out behavior in other Arctic pinnipeds
[@reder2003; @udevitz2009; @perry2017]. In particular, we linked the weighted
average daily locations from our bio-loggers to key weather values from the
North American Regional Reanalysis (NARR) model produced by the National Centers
for Environmental Prediction [@mesinger2006]. The NARR model assimilates
observational data to produce a long-term picture of weather over North America.
Numerous weather variables are made available across the region 8 times daily.
For this study, NARR weather values were subset to the extent of our study area
over the Bering and Chukchi seas at 3-hr intervals based on the native grid
resolution of 32 km (1024 sq. km). The following meteorological variables were
interpolated and assigned to seal locations using a bilinear method: 1) air
temperature at 2m above the earth's surface, 2) wind consisting of northerly and
easterly vector components converted to wind speed using the Euclidean norm, 3)
barometric pressure at sea level, and 4) precipitation (Table
\@ref(tab:tblCovariates)).

Models for all species include the following effects: day-of-year, solar hour,
temperature, wind speed, barometric pressure, precipitation, wind chill
(represented by a *wind:temperature* interaction [@udevitz2009]), and
day-of-year and time-of-day (solar hour) interactions to permit diurnal patterns
to change from March to July. Ribbon and spotted seal models included age-sex
class and interactions between day-of-year and age-sex class, but we omitted
these from bearded seal models due to poor representation of age-sex classes
(Table \@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect
(and an interaction with day-of-year), since bearded seals occupy a substantial
range and we were interested in possible differences in the timing of haul-out
behavior along a latitudinal gradient. We omitted the latitudinal effect from
ribbon and spotted seal models since, during the spring, these species are most
prevalent near the southern ice edge in the Bering Sea [@conn2014a].

Notably missing from this list of explanatory variables is any spatial-temporal
representation of sea ice concentration, area, or extent. This may seem
counterintuitive when modeling the haul-out behavior of seal species with such a
close association to sea ice; seals haul out in the presence of sea ice, and we
could assess the local concentration of sea ice during these events (see
[@olnes2020]). However, a major focus of this study is to develop models
applicable for aerial survey correction factors and using sea ice as a covariate
would almost certainly bias haul-out predictions towards those seals that are on
or near ice and therefore more likely to haul out. Since aerial surveys can only
detect seals on ice, abundance estimates would be missing a correction for those
seals that are away from ice (e.g., on foraging trips). Additionally, many of
the deployments consisted of a single device attached to the rear flipper of the
seal which do not provide at-sea locations, limiting our ability to fully
evaluate habitat preferences related to sea-ice. It should be noted that the
aerial survey counts are modeled with sea-ice and other habitat parameters. So,
any final abundance estimates that combine models of survey counts and our model
for proportion of seals hauled out on ice do account for the strong association
between these species and sea-ice in the Bering and Chukchi seas. Lastly, our
study was limited to the spring season when seal haul-out tendencies are
strongly influenced by pupping, nursing, breeding behavior, and molt and these
drivers are likely more influential than specific sea-ice concentration.
Crawford et al [-@crawford2019] compared haul-out probability models for ringed
seals and found those that only included season (and not sea-ice concentration)
were the most parsimonious. Further, studies to evaluate habitat selection in
relation sea-ice parameters and other environmental covariates are in progress.
For these reasons, we have elected not to use sea ice concentration as a
predictor for haul-out probability.

We assessed whether the annual variation in maximum spring sea ice extent in the
Bering Sea influences the seasonal peak of seal haul-out behavior. In
particular, we used sea ice concentration data from the Nimbus-7 SMMR and DMSP
SSM/I-SSMIS Passive Microwave Dataset, Version 1 [@cavalieri1996] to calculate
maximum sea ice extent. All sea ice concentration grid cells (25 km^2^) in the
study area with greater than 15% concentration were counted daily to get the total sea
ice extent for each day between 15 February and 15 July across all years.
Maximum spring sea ice extent was simply the largest daily count of grid cells
with greater than 15 percent concentration for each year.

@@

We followed Ver Hoef et al. [-@verhoef2014a] in using linear, quadratic, and
cubic effects of day-of-year to represent temporal changes in behavior. However,
unlike previous models for harbor seals [@london2012] and ice-associated seals
[@verhoef2014a], which treated hour-of-day as a 24-level categorical variable to
capture diurnal cycles, we adopted a continuous formulation based on Fourier
series that provides a flexible model while preserving the inherent circularity
needed for time-of-day effects (i.e., hour 0 should be equal to hour 24). It
also represents hour-of-day with 6 parameters, which is a considerable reduction
when compared to a 24 parameter variable, especially when fitting models with
interactions between hour-of-day and other variables (e.g., age-sex class,
day-of-year). According to this approach, we used the following specification
for hour-of-day effects:
